---
title: "Data Wrangling with `dplyr`"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(nycflights13)
library(tidyverse)
coffee <- read_csv("coffee.csv")
```


## Welcome

This tutorial addresses some of the basics of data wrangling with the package `dplyr`. 

In this tutorial, you will learn how to:
  
- Filter data
- Select unique rows
- Randomly select rows
- Select variables on a variety of conditions
- Group data
- Use the pipeline
- And more!
  

## Subset Observations: Filtering 

### Example 1


To filter data using `dplyr`, use the `filter()` command with the dataset and the conditions you are filtering on:

```{r filter-weather-ex1, exercise=TRUE, exercise.eval=TRUE}
filter(weather, wind_gust > 50)
```
This shows all of the flights that had a wind gust speed of over 50 miles per hour.

### Example 2


To filter on more than one condition, put an `&` between each condition:

```{r filter-weather-ex2, exercise=TRUE, exercise.eval=TRUE}
filter(weather, wind_gust > 35 & temp > 65)
```
This shows all of the flights that had wind gusts over 35 miles per hour *and* an average temperature of over 65 degress fahrenheit.

### Example 3


To filter on more than one category, we use the `%in%` operator:

```{r filter-flights-ex1, exercise=TRUE, exercise.eval=TRUE}
filter(flights, carrier %in% c("B6", "UA"))
```

### Exercise 1


Your turn! Filter the `flights` data where `carrier` is `B6` (JetBlue) and `dep_delay` is above or equal to 300 minutes.

```{r filter-flights-e1, exercise=TRUE}

```

```{r filter-flights-e1-solution}
filter(flights, carrier=="B6" & dep_delay >= 300)
```

### Exercise 2

What was the date of the flight with its origin at LGA, by UA carrier (United Airlines), and an air time of over 300 minutes?

```{r filter-quiz-flights, exercise=TRUE}

```

```{r filter-quiz-flights-hint}
filter(flights, carrier=="UA" & origin =="LGA", air_time > 300)
```


```{r filtering-quiz}
quiz(
  question("What was the date of the flight with origin LGA, by UA carrier, and an air time of over 300 minutes?",
    answer("09/14/2013"),
    answer("09/10/2013", correct = TRUE),
    answer("10/09/2013"),
    answer("12/09/2013")
  )
)
```



## Other Subset Observations

### Example 1: Distinct Observations


To see all of the potential answers for a dataset, use the `distinct` command:

```{r distinct-coffee, exercise=TRUE, exercise.eval=TRUE}
distinct(coffee)
```
Here you can see all of the distinct options from participants. In the actual dataset, there are over fifty-thousand entries but only ten distinct possibilities! 

### Example 2: Select a specific number of rows

a. To randomly select on a fraction of rows, use the `sample_frac()` command:

```{r sample-frac-coffee, exercise=TRUE, exercise.eval=TRUE}
sample_frac(coffee, size = 0.01, replace = TRUE)
```
This command resamples the coffee data, based on the set proportion. Using `replace = TRUE` allows for the same row to be used more than once.


b. We can randomly select a set number of rows using the `sample_n()` command:

```{r sample-n-coffee, exercise=TRUE, exercise.eval=TRUE}
sample_n(coffee, 15, replace = TRUE)
```


c. To select rows by position, we use the `slice()` command:

```{r slice-coffee, exercise=TRUE, exercise.eval=TRUE}
slice(coffee, 45000:45010)
```


### Example 3: Select and order entries

To select and order entries, use the `top_n()` command:

```{r top-coffee, exercise=TRUE, exercise.eval=TRUE}
top_n(flights, 3, air_time)
```
The top three longest air times are, in no particular order, 691, 686, and 695 minutes. Due to a tie of 686 minutes, there are four entries because the command will still show you the three top times, regardless of a tie.

## Subset Variables: Selecting

### Example 1

a. To select the columns where the variable name contains a specific character string, the command goes as follows:

```{r contains-str, exercise=TRUE, exercise.eval=TRUE}
select(flights, contains("dep"))
```
This shows all of the columns where the variable names have `dep` in it, which includes: dep_time, sched_dep_time, and dep_delay, so that you can see all of the information about the departure and its potential delays.

b. For selecting columns where the name ends with a character string:

```{r ends-with, exercise=TRUE, exercise.eval=TRUE}
select(flights, ends_with("time"))
```
Rather than showing all variable names that have `time` in it, we are only given the variables that *end* with `time`, which includes the columns for: dep_time, sched_dep_time, arr_time, sched_arr_time, and air_time, so that you can see all of the information about arrival, departure, and air time!

**NOTE:** You can also select all of the columns where the variable name starts with a character string by using `select(dataset, starts_with("nameStart"))`


c. To select all columns, we use `select(dataset, everything())`

This can be useful to get a snapshot of the columns you're working with.

d. For selecting columns where the name matches a regular expression, you use:

```{r matches, exercise=TRUE, exercise.eval=TRUE}
select(flights, matches(".t."))
```

```{r matches-quiz}
quiz(
  question("What does the `.` mean in the command above?",
    answer("The `.` means that the column should not start nor end with `t`", correct=TRUE),
    answer("The `.` means that the column should contain `.` surrounding the `t` in the variable name(s)"),
    answer("The `.` means that there should be at least one `.` next to the `t`")
  )
)
```

e. To select columns between two specific variables, you use:

```{r between-two, exercise=TRUE, exercise.eval=TRUE}
select(flights, arr_time:arr_delay)
```

f. For selecting all columns expect that of a specific variable, you use:

```{r all-but-1, exercise=TRUE, exercise.eval=TRUE}
select(flights, -arr_time)
```

**NOTE:** To select multiple columns to exclude, you can use `select(dataset, -c(column1, column4, column6))`


### Quiz

```{r select-quiz}
quiz(
  question("What command would you use if you wanted to select all of the dataset?",
    answer("select(dataset, contains(everything))"),
    answer("select(dataset, matches(everything))"),
    answer("select(dataset, everything())", correct=TRUE),
    answer("select(dataset, everything(all))")
  ),
  question("What command should you use if you know only one word in a variable name but don't know where it is in the name?",
    answer("select(dataset, contains('word'))", correct=TRUE),
    answer("select(dataset, matches('.word.'))"),
    answer("select(dataset, starts_with('word'))"),
    answer("select(dataset, ends_with('word'))")
  ),
  question("What command should you use if you want to get rid of three columns?",
    answer("select(dataset, -column1, -column2, -column3)"),
    answer("select(dataset, -column1, column2, column3)"),       
    answer("select(dataset, -c(column1, column2, column3))", correct=TRUE),
    answer("select(dataset, c(-column1, -column2, -column3))"))
)
```

## Using the Pipeline

### How & When to use the Pipeline

We use the pipeline, `%>%`, to get information in an organized manner, where we can get a filter, summarize, and select all in one chunk of code, without going through the hassle of performing each function separately. 

The pipeline turns this (code explained in next section):

```{r pipeline-1, exercise=TRUE, exercise.eval=TRUE}
select(mutate(flights, delay = arr_time - sched_arr_time), delay)
```

Into this: 

```{r mutate-1, exercise=TRUE, exercise.eval=TRUE}
flights %>%
  mutate(delay= arr_time-sched_arr_time) %>%
  select(delay)
```
While it may not feel like a significant difference when you're only using two functions, the pipeline really makes a difference when you're working with several functions, which we will see later on. 

## Making and Grouping Data

### Example 1

In order to compute and append one or more columns, we use the mutate command as follows:

```{r pipe-mutate, exercise=TRUE, exercise.eval=TRUE}
flights %>%
  mutate(delay= arr_time - sched_arr_time) %>%
  select(delay, arr_time, sched_arr_time)
```
This shows the arrival time delay, which would be more relevant if it wasn't already a column but now you know how to find this number!


### Example 2

In order to compute and append one or more new columns, while getting rid of all original columns:

```{r transmute, exercise=TRUE, exercise.eval=TRUE}
transmute(flights, delay = arr_time - sched_arr_time)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(nycflights13)
```


### Example 3

To group data into rows with the same value of a variable, you use:

```{r group-by1, exercise=TRUE, exercise.eval=TRUE}
flights %>%
  group_by(carrier)
```
You can also ungroup the rows by using `ungroup(flights)`.

a. Grouping is useful for when you want to create a separate summary for each group:

```{r group-summary, exercise=TRUE, exercise.eval=TRUE}
flights %>%
  group_by(carrier) %>% 
  summarise(mean(dep_delay, na.rm=TRUE))
```
This shows the average departure delay based on the flight carrier. The `na.rm=TRUE` removes all of the NAs from the calculation.

b. Grouping can also be used for mutating columns based on groups:

```{r group-mutate, exercise=TRUE, exercise.eval=TRUE}
flights %>%
  group_by(carrier, origin) %>%
  mutate(avg_delay = mean(dep_delay, na.rm=TRUE)) %>%
  select(carrier, origin, avg_delay) %>% 
  distinct()
```
This shows the average departure delay based on both the flight carrier and the airport origin.

### Exercise 1

Using the pipeline, find the mean departure delay (`dep_delay`) and mean arrival delay (`arr_delay`) and group by month to see if the time of year influences the delays. Be sure to only show each month once!

```{r pipe-filter-ex1, exercise=TRUE}

```

```{r pipe-filter-ex1-hint}
#remember to use `na.rm=TRUE` in the mean calculations

#use the `distinct()` command to output only one row for each month
```

```{r pipe-filter-ex1-solution}
flights %>%
  group_by(month) %>%
  mutate(mean_delay = mean(dep_delay, na.rm=TRUE), mean_arr = mean(arr_delay, na.rm=TRUE)) %>%
  select(month, mean_delay, mean_arr) %>% 
  distinct()
```

### Exercise 2

Using the pipeline, find the longest five delays by carrier. 

```{r top-n-ex, exercise=TRUE}

```

```{r top-n-ex-hint}
#use summarize(max_delay = max(dep_delay, na.rm=TRUE)) to calculate the max delays
```

```{r top-n-ex-solution}
flights %>%
  group_by(carrier) %>%
  summarize(max_delay = max(dep_delay, na.rm=TRUE)) %>% 
  top_n(5)
```






